\vspace{1.0cm} \LARGE \bf Revealing the competitive dynamics of influenza viruses through evolutionary cartography

### 2012-11-30 flux.tex
Revise results and include new figures.

Before 2009, there were four major clades of influenza circulating within the human population: A/H3N2, A/H1N1, B/Victoria and B/Yamagata. 
In the case of influenza A, subtypes refer to the genes, hemagglutinin (H or HA) and neuraminidase (N or NA), that are primarily responsible for the antigenic character of a strain. 
In the case of influenza B, lineages Victoria and Yamagata diverged from one prior to 1980 \cite{Rota92}.
Mutations to the HA1 region of the hemagglutinin protein are thought to drive the majority of antigenic drift in the influenza virus \cite{Nelson07NatRevGenet}. 
Experimental characterization of antigenic phenotype is possible through the hemagglutination inhibition (HI) assay \cite{Hirst43}, which measures the cross-reactivity of one virus strain to serum raised against another strain through challenge or vaccination. 
Sera from older strains react poorly with more evolved viruses resulting in new strains having a transmission advantage over established strains.
HI assays have shown there to be little to no cross-reactivity between influenza clades A/H3N2, A/H1N1, B/Victoria and B/Yamagata \cite{Hay01}.

@@

Previously, the antigenic and genetic patterns of influenza evolution have been analyzed essentially in isolation. 
An antigenic map is constructed from a panel of HI measurements, and a phylogenetic tree is constructed from sequence data. 
However, the opportunity for a combined approach exists as both the antigenic map and the phylogenetic tree often contain many of the same isolates. 
Here, we implement a flexible Bayesian approach to jointly characterize the antigenic and genetic evolution of the influenza virus population. 
We apply this approach to investigate the dynamics of clades A/H3N2, A/H1N1, B/Victoria and B/Yamagata.

@@

\subsection*{Antigenic cartography}

In order to assess patterns of antigenic evolution among influenza strains, we implemented a Bayesian probabilistic analog of multidimensional scaling, referred to here as BMDS (see Methods).
In this model, viruses and sera are given $N$-dimensional locations, thus specifying an `antigenic map', such that distances between viruses and sera in this space are inversely proportional to cross-reactivity.
In the BMDS model, a map distance of one antigenic unit translates to an expectation of a 2-fold drop in HI titer between virus and sera.
Maps that produce pairwise distances most congruent with the observed titers will have a high likelihood and will be favored by the BMDS model.
We integrate over sources of uncertainty, such as antigenic locations, in a flexible Bayesian fashion.
We apply this model to HI measurements of virus isolates against post-infection ferret sera for influenza A/H3N2, A/H1N1, B/Victoria and B/Yamagata (see Methods).

We test model performance by constructing a training dataset representing 90\% of the HI measurements of the A/H3N2 dataset and a test dataset representing the remaining 10\% of the full dataset. 
By fitting the BMDS model to the training dataset, we are able to predict HI titers in the test dataset and compare predicted titers to observed titers in the test dataset. 
We find that a BMDS analog of the model used by Smith et al.\ \cite{Smith04}, in which viruses and sera are represented as 2D locations and expected titer is relative to the maximum titer of a particular ferret sera, performs well, yielding an average absolute predictive error of 0.87 log$_2$ HI titers (Table \ref{errortable}).
We find that a model with two dimensional antigenic locations predicts test data to the same precision as a 3D model, but performs substantially better than a 1D model or models with four or more dimensions (Table \ref{errortable}), and thus specify a two dimensional model in all subsequent analyses.
We extend this model by estimating the strength of overall reactivity of each serum rather than fixing this at the maximum titer, and additionally, by estimating the strength of reactivity of each virus isolate.
We refer to these estimates as serum effects and virus effects, respectively.
We found that including these effects decreased test error to 0.75 for influenza A/H3N2.
We arrive at similar precisions when repeating this analysis for A/H1N1, B/Victoria and B/Yamagata.
For the 2D model with serum and virus effects, A/H1N1 shows a test error of 0.64, B/Victoria shows a test error of 0.74 and B/Yamagata shows a test error of 0.80.

@@

\subsection*{Antigenic drift across influenza lineages}

%%% dynamicstable %%%
\begin{table}[tb]
	\centering
	\caption{\textbf{Antigenic dynamics of influenza lineages.}}
	\label{dynamicstable}
	\begin{tabular}{ l c c c c } 
	\hline
	 													&	A/H3N2	& 	A/H1N1 	& 	B/Vic	& 	B/Yam	\\
	\hline	
	Rate of population drift 							&	1.004	&	0.554	&	0.406	&	0.205	\\
	Population standard deviation 						&	0.787	&	0.790	&	0.917	&	0.631	\\

	Rate of drift on side branches						&	0.067	&	0.192	&	-0.024	&	0.025	\\	
	Rate of drift on internal branches					&	0.172	&	0.017	&	0.035	&	0.009	\\
	Rate of drift on trunk								&	1.011	&	0.465	&	0.015	&	0.417	\\

	Coefficient of diffusion on side branches 			&	0.425	&	0.387	&	0.461	&	0.251	\\
	Coefficient of diffusion on internal branches		&	0.290	&	0.161	&	0.259	&	0.154	\\
	Coefficient of diffusion on trunk  					&	0.511	&	0.442 	&	1.098	&	0.260	\\	
	\hline
	\end{tabular}
\end{table}

Through this analysis we find that the antigenic phenotype of influenza A (H3N2) underwent rapid turnover from 1968 to 2011 (Figure~\ref{flu_grid}A).  
Here, cross-reactive measurements only exist betwen strains sampled at most 11 years apart, leaving only threshold titers, e.g.\ `$<$40', in more temporally distant comparisons.  
Because of the threshold of sensitivity of the HI assay, it's impossible to distinguish a linear trajectory in 2D antigenic space, from a curved trajectory, so long as the curve does not bring antigenic phenotype full circle to have cross-reactive measurements between temporally distant strains.
To solve this problem of identifiability, we assumed a weak prior that favors linear movement in the 2D antigenic space, with the slope of the linear relationship and the precision of the relationship incorporated into the Bayesian model (see Methods).

We find that influenza A (H3N2) evolved 1.16 antigenic units per year from 1968 to 2011, with 95\% highest posterior density (HPD) interval 1.13--1.20 (Figure~\ref{flu_grid}C).
However, the rate of antigenic drift was not constant, with year-to-year movement of mean antigenic phenotype showing an interquartile range of 0.40--1.79.  
Large jumps in antigenic phenotype (Figure~\ref{flu_grid}C) correspond to cluster transitions identified by Smith et al. \cite{Smith04}.  
Most variation is contained within the first antigenic dimension, but dimension 2 ocassionally shows variation when two antigenically distinct lineages emerge and transiently coexist (Figure~\ref{flu_grid}E), as is the case with the previously identified Beijing 1989 and Beijing 1992 clusters.

@@

%%% flu_grid %%%
\begin{figure}[tb]
	\centering		
	\includegraphics[width=\textwidth]{figures/flu_grid}
	\caption{\textbf{Antigenic locations of influenza H3N2 and H1N1.} 
	(A) and (B) Antigenic maps showing the mean posterior location of 338 strains of H3N2 influenza and 243 strains of H1N1 influenza.  
	The model has been constructed so that the primary axis of variation lies along the $x$-axis (AG1), with the $y$-axis (AG2) orthogonal to this axis.  
	(C) and (D) Antigenic location along the primary axis of variation (AG1) vs.\ year of virus isolation.  
	The dashed lines show the relationship of between time and AG1 with a slope of \tbc{XXX} for H3N2 and \tbc{XXX} for H1N1.  
	(E) and (F) Antigenic location along the secondary axis of variation (AG2) vs.\ year of virus isolation.  
	Points represent median strain locations and translucent areas represent 50\% highest density regions.
	Antigenic units represent two-fold dilutions of the HI assay, and strains have been colored based on year of isolation.} 
	\label{flu_grid} 
\end{figure}

@@

We investigated the relative rates of antigenic evolution in different influenza lineages from 1999 to 2009, finding a general correspondence between antigenic drift along antigenic dimension 1 and relative incidence between influenza A/H3N2, A/H1N1, B/Victoria and B/Yamagata.
From 1999 to 2009, influenza A (H3N2) accounted for the majority of antigenic evolution (51\%), while A/H1N1, B/Victoria and B/Yamagata split the remainder, accounting for 17\%, 18\% and 14\%, respectively.
Similarly, influenza A (H3N2) was responsible for the majority of incidence (54\%) in the USA during this time period.
Influenza A (H1N1) accounted for 23\% of incidence, B/Victoria accounted for 12\% of incidence and B/Yamagata accounted for 10\% of incidence.
These proportions are significantly similar to one another (randomization test $p = 0.010$).

%%% drift_incidence %%%
\begin{figure}[tb]
	\centering		
	\includegraphics[width=\textwidth]{figures/drift_incidence}
	\caption{\textbf{Relative antigenic drift and incidence between lineages of influenza from 1999 to 2009.} 
	(A) Relative antigenic drift of influenza A H3N2 and H1N1 and influenza B Victoria and Yamagata lineages.
	Relative change in antigenic dimension 1 between year $i$ and year $i-1$ is shown for each of the 4 lineages.
	(B) Relative incidence of influenza A H3N2 and H1N1 and influenza B Victoria and Yamagata lineages.
	Relative isolation counts of each lineage in the USA for each winter influenza season.
	} 
	\label{drift_incidence} 
\end{figure}

%%% DISCUSSION %%%
\section*{Discussion}

@@

Let $X_i \in \Re^{P}$ represent the cartographic location of virus $i$ for $i = 1,\ldots,\vn$, and $Y_j$ represent the cartographic location of serum $j$ for $j = 1,\ldots,\sn$.
Virus and antiserum may be isolated from / raised against the same strain and have different cartographic locations.
Typically, $P = 2$, but higher or lower dimensions may better reflect the data.  
This gives set of distances between virus and serum cartographic locations 
\begin{equation}
	\delta_{ij} =  || X_i - Y_j ||_2.
\end{equation}
Here, $H_{ij}$ represents the log$_2$ titer of virus $i$ against serum $j$, and immunological distance can be defined as
\begin{equation}
	d_{ij} =  S_j - H_{ij},
\end{equation}
where $S_j = \max ( H_{1j},\ldots,H_{\vn j} )$.
Traditionally, the goal of multidimensional scaling (MDS) optimizes over $\mathbf{X}$ and $\mathbf{Y}$ such that
\begin{equation}
	\sum_{(i,j) \in \cal I} 
	\left(
		\delta_{ij} - d_{ij}
	\right)^2
\end{equation}
is minimized, where $\cal I = \{ (i,j) : H_{ij} \mbox{ is measured} \}$. 
Here, we instead assume a probabilistic interpretation in which an observed titer is normally distributed around its cartographic expectation
\begin{equation} 
	H_{ij} \sim \mbox{Normal}( S_j - \delta_{ij}, \mdssd^2 ),
\end{equation}
and the likelihood of observing a particular titer given the placement of antigenic locations is 
\begin{equation} 
	\point(H_{ij}) = \phi \left( \frac{ H_{ij} + \delta_{ij} - S_j }{ \mdssd } \right),
\end{equation}
where $\mdssd^2$ represents variance and $\phi(\cdot)$ represents the standard normal PDF.
HI assays sometimes show no inhibition at all measured titrations, e.g.\ a measurement can be reported as `$<$40'.
In this case, the likelihood of observing the threshold measurement follows the cumulative density of the lower tail of the normal distribution
\begin{equation} 
	\threshold(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - S_j }{ \mdssd } \right),
\end{equation}
where $\Phi(\cdot)$ represents the standard normal CDF.
Although it is simplest to assume that immunological measurements represent point estimates, it seems more natural to assume that the threshold for inhibition occurs between two titers, e.g.\ we observe inhibition at 1:40 dilution and no inhibition at 1:80 dilution.
Rather than taking the HI titer as 40, we can instead treat this as an interval measurement, assuming that the exact titer for inhibition would occur somewhere between 40 and 80.
HI titers are usually reported as the highest titer that successfully inhibits virus binding or growth, so that in this case, we calculate the likelihood of an interval measurement as
\begin{equation} 
	\interval(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - S_j + 1 }{ \mdssd } \right) - \Phi \left( \frac{ H_{ij} + \delta_{ij} - S_j }{\mdssd} \right).
\end{equation}
These likelihoods are illustrated in Figure \ref{hij_likelihood}.
Throughout our analyses, we use interval likelihoods $\interval$ rather than point likelihoods $\point$ unless otherwise noted.

@@

As presented, virus and serum locations $\mathbf{X}$ and $\mathbf{Y}$ accurately characterize antigenic distances between virus strains based on titers from pairwise HI assays.
However, these distances represent only a local description of antigenic space and do not provide a global picture.
An example of this phenomena is shown in Figure \ref{schematic_map}.
In this case it is impossible to determine from the HI data at hand whether the blue and yellow viruses are antigenically similar or antigenically divergent. 
This presents an issue of model identifiability; multiple antigenic maps give the same likelihood of observing the serological data.
In order to achieve more interpretable antigenic locations we impose a weak prior on global locations.
In influenza, it's clear that antigenic distance between strains increases with time \cite{Smith04,Cai10}, and so we impose a prior in which virus locations are distributed according to a multivariate normal distribution in which the expectation of the first antigenic dimension increases according to date of virus sampling
\begin{equation} 
	X_{i} \sim \mbox{Normal}( \{\beta \, y_i, 0\}, \Sigma )
\end{equation}
for $P=2$, where $y_i$ is the difference in time between virus $i$ and the earliest sampled virus and covariance matrix $\Sigma$ is equal to $\driftsd^2$ along diagonal entries and equal to zero along off-diagonal entries.
Thus, this model assumes that the virus population drifts in a line across the antigenic map at rate $\beta$.
The parameter $\driftsd$ determines the breadth of the virus population at each point in time.
We assume that both the prior slope $\beta$ and the prior precision $1/\driftsd^2$ follow a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=1000$.
We further introduce Metropolis-Hastings proposals that scale $\beta$ and $\driftsd$.

@@

Most previous work on virus antigenic and genetic evolution has reconstructed antigenic and genetic patterns in isolation, and then compared the results of each reconstruction to assess congruence of these two aspects of evolution \cite{Hay01,Smith04,Russell08}.  
Here, we simultaneously model antigenic locations and genetic relatedness by assuming that viruses diffuse across the antigenic map according to a Brownian motion process \cite{Lemey10}.
We assume a prior on virus locations according to the evolutionary process and leave a flat prior on serum locations
\begin{eqnarray}
	(X_1,\ldots,X_n) &\sim& \mbox{Evolutionary Brownian Process}(\tree, \diffusionsd) \nonumber \\
	Y_j &\sim& \mathrm{Uniform}( -\infty, \infty ),
\end{eqnarray}
where $\tree$ is a phylogeny specifying tree topology and branch lengths and $\diffusionsd$ is the volatility parameter of the Brownian motion,  set to have equal variance in each antigenic dimension and no correlation across dimensions.
Thus, viruses which are genetically similar are induced to have prior locations close to one another on the antigenic map.
The phylogenetic tree $\tree$ is estimated using sequence data for viruses $1,\ldots,\vn$ according to well established methods implemented in the software package BEAST \cite{BEAST}.
The volatility parameter $\diffusionsd$ is estimated jointly from the genetic and antigenic data.
Parameter $\diffusionsd$ is assumed to follow a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=1000$.
Metropolis-Hastings proposals scale $\diffusionsd$.
The probability of observing virus locations $p(\mathbf{X}|\tree,\diffusionsd)$ is determined through analytical integration across internal states following the methods introduced in \cite{Lemey10}.
Thus, under the full model, the posterior probability of observing virus and serum locations given immunological data and phylogeny is factored
\begin{equation}
	p(\mathbf{X},\mathbf{Y} | \mathbf{H},\tree) = \frac{ p(\mathbf{H}|\mathbf{X},\mathbf{Y},\mathbf{S},\mathbf{V}, \mdssd) \; 
	p(\mathbf{X} | \beta, \vartheta) \;
	p(\mathbf{X} | \tree, \diffusionsd) \; 
	p(\mathbf{Y},\mathbf{S},\mathbf{V},\mdssd,\beta,\vartheta,\diffusionsd)}{ p(\mathbf{H}) }.
\end{equation}

>>>>

Before 2009, there were four major clades of influenza circulating within the human population: influenza A H3N2 and H1N1, and influenza B Victoria and Yamagata. 
In the case of influenza A, subtypes A/H3N2 and A/H1N1 refer to the genes, hemagglutinin (H or HA) and neuraminidase (N or NA), that are primarily responsible for the antigenic character of a strain. 
In the case of influenza B, B/Vic and B/Yam refer to antigenically distinct lineages which diverged from one prior to 1980 \cite{Rota92}.
Mutations to the HA1 region of the hemagglutinin protein are thought to drive the majority of antigenic drift in the influenza virus \cite{Nelson07NatRevGenet}. 
Experimental characterization of antigenic phenotype is possible through the hemagglutination inhibition (HI) assay \cite{Hirst43}, which measures the cross-reactivity of one virus strain to serum raised against another strain through challenge or vaccination. 
Sera from older strains react poorly with more evolved viruses resulting in new strains having a transmission advantage over established strains.
HI assays have shown there to be little to no cross-reactivity between influenza clades A/H3N2, A/H1N1, B/Vic and B/Yam \cite{Hay01}.

@@

Previously, the antigenic and genetic patterns of influenza evolution have been analyzed essentially in isolation. 
An antigenic map is constructed from a panel of HI measurements, and a phylogenetic tree is constructed from sequence data. 
However, the opportunity for a combined approach exists as both the antigenic map and the phylogenetic tree often contain many of the same isolates. 
Here, we implement a flexible Bayesian approach to jointly characterize the antigenic and genetic evolution of the influenza virus population. 
We apply this approach to investigate the dynamics of clades A/H3N2, A/H1N1, B/Vic and B/Yam.

@@

\subsection*{Antigenic and evolutionary cartography}

In order to assess patterns of antigenic evolution among influenza strains, we implemented a Bayesian probabilistic analog of multidimensional scaling, referred to here as BMDS (see Methods).
In this model, viruses and sera are given $N$-dimensional locations, thus specifying an `antigenic map', such that distances between viruses and sera in this space are inversely proportional to cross-reactivity.
In the BMDS model, a map distance of one antigenic unit translates to an expectation of a 2-fold drop in HI titer between virus and sera.
Maps that produce pairwise distances most congruent with the observed titers will have a high likelihood and will be favored by the BMDS model.
We integrate over sources of uncertainty, such as antigenic locations, in a flexible Bayesian fashion.
We apply this model to HI measurements of virus isolates against post-infection ferret sera for influenza A/H3N2, A/H1N1, B/Vic and B/Yam (see Methods).

We test model performance by constructing a training dataset representing 90\% of the HI measurements of the A/H3N2 dataset and a test dataset representing the remaining 10\% of the full dataset. 
By fitting the BMDS model to the training dataset, we are able to predict HI titers in the test dataset and compare predicted titers to observed titers in the test dataset. 
We find that a BMDS analog of the model used by Smith et al.\ \cite{Smith04}, in which viruses and sera are represented as 2D locations and expected titer is relative to the maximum titer of a particular ferret sera, performs well, yielding an average absolute predictive error of 0.87 log$_2$ HI titers (Table \ref{errortable}).
We find that a model with two dimensional antigenic locations predicts test data to the same precision as a 3D model, but performs substantially better than a 1D model or models with four or more dimensions (Table \ref{errortable}), and thus specify a two dimensional model in all subsequent analyses.
We extend this model by estimating the strength of overall reactivity of each serum rather than fixing this at the maximum titer, and additionally, by estimating the strength of reactivity of each virus isolate.
We refer to these estimates as serum effects and virus effects, respectively.
We found that including these effects decreased test error to 0.75 for influenza A/H3N2.
We arrive at similar precisions when repeating this analysis for A/H1N1, B/Vic and B/Yam.
For the 2D model with serum and virus effects, A/H1N1 shows a test error of 0.64, B/Vic shows a test error of 0.74 and B/Yam shows a test error of 0.80.

@@

Most previous work on virus antigenic and genetic evolution has examined antigenic and genetic patterns in isolation, and then compared the results to assess congruence of these two aspects of evolution \cite{Hay01,Smith04,Russell08}. 
Here, we simultaneously model antigenic and genetic evolution by adapting an evolutionary diffusion process \cite{Lemey10}, wherein a virus's antigenic character state evolves along branches of phylogenetic tree according to Brownian motion process (see Methods).
Here, the diffusion process acts as a prior, so that genetically similar viruses are expected to share similar antigenic phenotypes, and we model the relationship between genetic change and accrual of antigenic distance.
We include sequence data for A/H3N2, A/H1N1, B/Vic and B/Yam to estimate this diffusion process.
We find similar levels of predictive accuracy when including the diffusion process with A/H3N2 showing a test error of 0.75, A/H1N1 showing a test error of 0.66, B/Vic showing a test error of 0.70 and B/Yam showing a test error of 0.78.

\subsection*{Antigenic evolution across influenza lineages}

Through our modeling, we reveal the evolutionary basis for antigenic phenotype in influenza A/H3N2, A/H1N1, B/Vic and B/Yam (Figure~\ref{map}).
We observe prominent antigenic clusters in A/H3N2 and A/H1N1 with less obvious clustering in B/Vic and B/Yam.
Antigenic clusters show high genetic similarity, so that we only observe a single diffusion event leading to each cluster, rather than the repeated emergence of clusters.
This analysis makes the fate of antigenic clusters obvious, with two clusters in A/H3N2 (Texas/77 and Beijing/89) appearing to be evolutionary dead-ends.

%%% map %%%
\begin{figure}[h]
	\centering		
	\includegraphics[width=0.85\textwidth]{figures/map}
	\caption{\textbf{Antigenic locations of A/H3N2, A/H1N1, B/Vic and B/Yam viruses showing evolutionary relationships between virus samples.} 
	Circles represent a posterior sample of virus locations and have been colored based on year of isolation in 4-year intervals.
	Antigenic units represent two-fold dilutions of the HI assay.
	Distances between clades, e.g.\ A/H3N2 and A/H1N1, are arbitrary.
	Lines represent mean posterior diffusion paths when virus locations are fixed.} 
	\label{map} 
\end{figure}

For A/H3N2, cross-reactive measurements only exist betwen strains sampled at most 11 years apart, leaving only threshold titers, e.g.\ `$<$40', in more temporally distant comparisons.  
Because of the threshold of sensitivity of the HI assay, it's impossible to distinguish a linear trajectory in 2D antigenic space, from a curved trajectory, so long as the curve does not bring antigenic phenotype full circle to have cross-reactive measurements between temporally distant strains.
To solve this problem of identifiability, we assumed a weak prior that favors linear movement in the 2D antigenic space, with the slope of the linear relationship and the precision of the relationship incorporated into the Bayesian model (see Methods).

We find that influenza A (H3N2) evolved 1.16 antigenic units per year from 1968 to 2011, with 95\% highest posterior density (HPD) interval 1.13--1.20 (Figure~\ref{drift}).
However, the rate of antigenic drift was not constant, with year-to-year movement of mean antigenic phenotype showing an interquartile range of 0.40--1.79.  
Large jumps in antigenic phenotype (Figure~\ref{drift}) correspond to cluster transitions identified by Smith et al. \cite{Smith04}.  
Most variation is contained within the first antigenic dimension, but dimension 2 ocassionally shows variation when two antigenically distinct lineages emerge and transiently coexist (Figure~\ref{map}), as is the case with the previously identified Beijing 1989 and Beijing 1992 clusters.

%%% drift %%%
\begin{figure}[h]
	\centering		
	\includegraphics[width=0.65\textwidth]{figures/drift}
	\caption{\textbf{Antigenic drift of A/H3N2, A/H1N1, B/Vic and B/Yam viruses showing evolutionary relationships between virus samples.} 
	Antigenic drift is shown in terms of change of location in the first antigenic dimension through time.
	Circles represent a posterior sample of virus locations and have been colored based on year of isolation in 4-year intervals.
	Antigenic units represent two-fold dilutions of the HI assay.
	Distances between clades, e.g.\ A/H3N2 and A/H1N1, are arbitrary.
	Lines represent mean posterior diffusion paths when virus locations are fixed.} 
	\label{drift} 
\end{figure}

@@

@@

We investigated the relative rates of antigenic evolution in different influenza lineages from 1999 to 2009, finding a general correspondence between antigenic drift along antigenic dimension 1 and relative incidence between influenza A/H3N2, A/H1N1, B/Vic and B/Yam.
From 1999 to 2009, influenza A (H3N2) accounted for the majority of antigenic evolution (51\%), while A/H1N1, B/Vic and B/Yam split the remainder, accounting for 17\%, 18\% and 14\%, respectively.
Similarly, influenza A (H3N2) was responsible for the majority of incidence (54\%) in the USA during this time period.
Influenza A (H1N1) accounted for 23\% of incidence, B/Vic accounted for 12\% of incidence and B/Yam accounted for 10\% of incidence.
These proportions are significantly similar to one another (randomization test $p = 0.010$).

%%% incidence %%%
\begin{figure}[tb]
	\centering		
	\includegraphics[width=\textwidth]{figures/incidence}
	\caption{\textbf{Relative antigenic drift and incidence between lineages of influenza from 1999 to 2009.} 
	(A) Relative antigenic drift of influenza A H3N2 and H1N1 and influenza B Victoria and Yamagata lineages.
	Relative change in antigenic dimension 1 between year $i$ and year $i-1$ is shown for each of the 4 lineages.
	(B) Relative incidence of influenza A H3N2 and H1N1 and influenza B Victoria and Yamagata lineages.
	Relative isolation counts of each lineage in the USA for each winter influenza season.
	} 
	\label{incidence} 
\end{figure}

@@

Let $X_i \in \Re^{P}$ represent the cartographic location of virus $i$ for $i = 1,\ldots,\vn$, and $Y_j$ represent the cartographic location of serum $j$ for $j = 1,\ldots,\sn$.
Virus and antiserum may be isolated from / raised against the same strain and have different cartographic locations.
Typically, $P = 2$, but higher or lower dimensions may better reflect the data.  
This gives set of distances between virus and serum cartographic locations 
\begin{equation}
	\delta_{ij} =  || X_i - Y_j ||_2.
\end{equation}
Here, $H_{ij}$ represents the log$_2$ titer of virus $i$ against serum $j$, and immunological distance can be defined as
\begin{equation}
	d_{ij} =  S_j - H_{ij},
\end{equation}
where $S_j = \max ( H_{1j},\ldots,H_{\vn j} )$.
Traditionally, the goal of multidimensional scaling (MDS) optimizes over $\mathbf{X}$ and $\mathbf{Y}$ such that
\begin{equation}
	\sum_{(i,j) \in \cal I} 
	\left(
		\delta_{ij} - d_{ij}
	\right)^2
\end{equation}
is minimized, where $\cal I = \{ (i,j) : H_{ij} \mbox{ is measured} \}$. 
Here, we instead assume a probabilistic interpretation in which an observed titer is normally distributed around its cartographic expectation
\begin{equation} 
	H_{ij} \sim \mbox{Normal}( S_j - \delta_{ij}, \mdssd^2 ),
\end{equation}
and the likelihood of observing a particular titer given the placement of antigenic locations is 
\begin{equation} 
	\point(H_{ij}) = \phi \left( \frac{ H_{ij} + \delta_{ij} - S_j }{ \mdssd } \right),
\end{equation}
where $\mdssd^2$ represents variance and $\phi(\cdot)$ represents the standard normal PDF.
HI assays sometimes show no inhibition at all measured titrations, e.g.\ a measurement can be reported as `$<$40'.
In this case, the likelihood of observing the threshold measurement follows the cumulative density of the lower tail of the normal distribution
\begin{equation} 
	\threshold(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - S_j }{ \mdssd } \right),
\end{equation}
where $\Phi(\cdot)$ represents the standard normal CDF.
Although it is simplest to assume that immunological measurements represent point estimates, it seems more natural to assume that the threshold for inhibition occurs between two titers, e.g.\ we observe inhibition at 1:40 dilution and no inhibition at 1:80 dilution.
Rather than taking the HI titer as 40, we can instead treat this as an interval measurement, assuming that the exact titer for inhibition would occur somewhere between 40 and 80.
HI titers are usually reported as the highest titer that successfully inhibits virus binding or growth, so that in this case, we calculate the likelihood of an interval measurement as
\begin{equation} 
	\interval(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - S_j + 1 }{ \mdssd } \right) - \Phi \left( \frac{ H_{ij} + \delta_{ij} - S_j }{\mdssd} \right).
\end{equation}
These likelihoods are illustrated in Figure~\ref{hij_likelihood}.
Throughout our analyses, we use interval likelihoods $\interval$ rather than point likelihoods $\point$ unless otherwise noted.

@@

As presented, virus and serum locations $\mathbf{X}$ and $\mathbf{Y}$ accurately characterize antigenic distances between virus strains based on titers from pairwise HI assays.
However, these distances represent only a local description of antigenic space and do not provide a global picture.
An example of this phenomena is shown in Figure~\ref{schematic_map}.
In this case it is impossible to determine from the HI data at hand whether the blue and yellow viruses are antigenically similar or antigenically divergent. 
This presents an issue of model identifiability; multiple antigenic maps give the same likelihood of observing the serological data.
In order to achieve more interpretable antigenic locations we impose a weak prior on global locations.
In influenza, it's clear that antigenic distance between strains increases with time \cite{Smith04,Cai10}, and so we impose a prior in which virus locations are distributed according to a multivariate normal distribution in which the expectation of the first antigenic dimension increases according to date of virus sampling
\begin{equation} 
	X_{i} \sim \mbox{Normal}( \{\beta \, y_i, 0\}, \Sigma )
\end{equation}
for $P=2$, where $y_i$ is the difference in time between virus $i$ and the earliest sampled virus and covariance matrix $\Sigma$ is equal to $\driftsd^2$ along diagonal entries and equal to zero along off-diagonal entries.
Thus, this model assumes that the virus population drifts in a line across the antigenic map at rate $\beta$.
The parameter $\driftsd$ determines the breadth of the virus population at each point in time.
We assume that both the prior slope $\beta$ and the prior precision $1/\driftsd^2$ follow a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=1000$.
We further introduce Metropolis-Hastings proposals that scale $\beta$ and $\driftsd$.

@@

We simultaneously model antigenic locations and genetic relatedness by assuming that viruses diffuse across the antigenic map according to a Brownian motion process \cite{Lemey10}.
We assume a prior on virus locations according to the evolutionary process and leave a flat prior on serum locations
\begin{eqnarray}
	(X_1,\ldots,X_n) &\sim& \mbox{Evolutionary Brownian Process}(\tree, \diffusionsd) \nonumber \\
	Y_j &\sim& \mathrm{Uniform}( -\infty, \infty ),
\end{eqnarray}
where $\tree$ is a phylogeny specifying tree topology and branch lengths and $\diffusionsd$ is the volatility parameter of the Brownian motion,  set to have equal variance in each antigenic dimension and no correlation across dimensions.
Thus, viruses which are genetically similar are induced to have prior locations close to one another on the antigenic map.
The phylogenetic tree $\tree$ is estimated using sequence data for viruses $1,\ldots,\vn$ according to well established methods implemented in the software package BEAST \cite{BEAST}.
The volatility parameter $\diffusionsd$ is estimated jointly from the genetic and antigenic data.
Parameter $\diffusionsd$ is assumed to follow a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=1000$.
Metropolis-Hastings proposals scale $\diffusionsd$.
The probability of observing virus locations $p(\mathbf{X}|\tree,\diffusionsd)$ is determined through analytical integration across internal states following the methods introduced in \cite{Lemey10}.
Thus, under the full model, the posterior probability of observing virus and serum locations given immunological data and phylogeny is factored
\begin{equation}
	p(\mathbf{X},\mathbf{Y} | \mathbf{H},\tree) = \frac{ p(\mathbf{H}|\mathbf{X},\mathbf{Y},\mathbf{S},\mathbf{V}, \mdssd) \; 
	p(\mathbf{X} | \beta, \vartheta) \;
	p(\mathbf{X} | \tree, \diffusionsd) \; 
	p(\mathbf{Y},\mathbf{S},\mathbf{V},\mdssd,\beta,\vartheta,\diffusionsd)}{ p(\mathbf{H}) }.
\end{equation}
